---
title: "population projection"
author: "Martin Monkman"
date: "2024-02-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)

library(here)

library(cansim)


## ---
## FUNCTIONS

# absolute value AND comma function
# from https://stackoverflow.com/questions/37949473/how-do-you-make-ggplot-labels-use-both-labels-abs-and-labels-comma
abs_comma <- function (x, ...) {
  format(abs(x), ..., big.mark = ",", scientific = FALSE, trim = TRUE)
}


```


## data

The Daily 2022-08-22
Population Projections for Canada, Provinces and Territories, 2021 to 2068, 2022

https://www150.statcan.gc.ca/n1/daily-quotidien/220822/dq220822b-eng.htm




Population Projections for Canada (2021 to 2068), Provinces and Territories (2021 to 2043)

https://www150.statcan.gc.ca/n1/pub/91-520-x/91-520-x2022001-eng.htm


All the tables
https://www150.statcan.gc.ca/n1/en/type/data?text=population+projection


Projected population, by projection scenario, age and sex, as of July 1 (x 1,000)1
Frequency: Annual

Table: 17-10-0057-01 (formerly CANSIM 052-0005)

Release date: 2022-08-22

Geography: Canada, Province or territory
https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1710005701


Population projections of the Canadian population, for Canada, provinces and territories (x 1,000)1
Frequency: Annual

Table: 17-10-0057-02

Release date: 2022-08-22

Geography: Canada, Province or territory
https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1710005702




```{r}

df_1710005701 <- get_cansim("17-10-0057-01")

df_1710005701 <- df_1710005701 |> 
  janitor::clean_names()

```

**NOTE**

Use "M1" medium-growth scenario


```{r}

df_1710005701_bc <- df_1710005701 |> 
  filter(
    geo == "British Columbia",
    projection_scenario == "Projection scenario M1: medium-growth",
    sex != "Both sexes"
    ) |> 
  filter(
    age_group %in% c(
      "0 to 4 years",
      "5 to 9 years",
      "10 to 14 years",
      "15 to 19 years",
      "20 to 24 years",
      "25 to 29 years",
      "30 to 34 years",
      "35 to 39 years",
      "40 to 44 years",
      "45 to 49 years",
      "50 to 54 years",
      "55 to 59 years",
      "60 to 64 years",
      "65 to 69 years",
      "70 to 74 years",
      "75 to 79 years",
      "80 to 84 years",
      "85 to 89 years",
      "90 to 94 years",
      "95 to 99 years",
      "100 years and over"
    )
  )

df_1710005701_bc

```


## data summary

```{r}

df_1710005701_bc |> 
  group_by(ref_date) |> 
  summarise(total_pop = sum(value))

```




```{r}

## population pyramid

# data
df_pop_pyr <- df_1710005701_bc

df_pop_pyr <- df_pop_pyr |> 
# merge LT1 + 1-4 into 0-4
#  mutate(`0 to 4` = LT1 + `1 to 4`) |> 
  select(
    geo,
    ref_date,
    sex,
    age_group,
    value
    )  

df_pop_pyr <- df_pop_pyr |> 
  mutate(age_grp = fct_drop(age_group))

levels(df_pop_pyr$age_grp)



# set male values to negative
df_pop_pyr <- df_pop_pyr |> 
  mutate(
    pop_total = case_when(
      sex == "Males" ~ value * -1000,
      sex == "Females" ~ value * 1000,
      TRUE ~ NA
    )
  )

df_pop_pyr

```


plot single year / region

```{r}
# set year and region
plot_year <- 2033

plot_region <- "British Columbia"
#plot_region <- "Vancouver Island/Coast"


# and plot
df_pop_pyr_1yr <- df_pop_pyr |> 
  filter(geo == plot_region) |> 
  filter(ref_date == plot_year)

df_pop_pyr_1yr

```


```{r}
pl_pyramid <-  
  ggplot(
    df_pop_pyr_1yr,
    aes(x = pop_total,
           y = age_grp,
           fill = sex
           )) + 
  geom_col() +
  theme_classic() +
  scale_fill_manual(
    values=c("#0C3343", "#EF5A27"),          # dark blue and action orange
    labels=c("men", "women")
    )  +
  xlab("population") +
  ylab(NULL)

pl_pyramid

# fix scale numbering
```


### plot labels



```{r}
pop_range_breaks <- seq(-300000, 300000, by = 100000)

pl_pyramid_total <- pl_pyramid +
  scale_x_continuous(
    breaks  = pop_range_breaks,
    #                      limits = c(min(pop_range_breaks), max(pop_range_breaks)),
#    limits = c(-1000, 1000),
#   labels = abs(pop_range_breaks)
#    labels = scales::comma
   labels = abs_comma  # custom function that combines absolute value AND commas
  )  +
#  scale_y_discrete(
#    breaks = c('0', '10', '20', '30', '40', '50', '60', '70', '80', '90+')
#  ) +
# add vertical white "gridlines"
  geom_vline(xintercept = pop_range_breaks, col="white") +
  labs(
    title = glue::glue("{plot_region}: {plot_year}"),
#       subtitle = glue::glue("total population = {community_total_population}"),

       caption = "Statistics Canada"
       ) +
  theme(legend.position = "bottom")

pl_pyramid_total


ggsave(here::here(glue::glue("plot_output/pyramid_{plot_region}_{plot_year}_population.jpg")))

```


