---
title: "cancensus_wrangling"
output: 
  html_notebook:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: inline
---

#`cancensus` - an introduction

** [`cancensus` github repo](https://github.com/mountainMath/cancensus) **


** [`cancensus` reference guide](https://mountainmath.github.io/cancensus/index.html)





```{r Setup}

# tidyverse
library(tidyverse)
library(here)

# cancensus

# (This is not necessary - `cancensus` is on CRAN now!)
#devtools::install_github("mountainMath/cancensus")

library(cancensus)


# set working directory
options(cancensus.cache_path = here("cancensus", "cache"))

```



```{r}

# set API key (stored in text file "secretAPI.txt")
secretAPI <- readLines(here("cancensus", "secretAPI.txt"))

options(cancensus.api_key = secretAPI)

```


***

Start exploring the data
(straight from [the `cancensus` reference guide](https://mountainmath.github.io/cancensus/index.html))




```{r}

# To view available Census datasets
census_datasets <- list_census_datasets()
census_datasets

# To view available named regions at different levels of Census hierarchy for the 2016 Census (for example)
census_regions16 <- list_census_regions("CA16")
census_regions21 <- list_census_regions("CA16")

# To view available Census variables
census_vectors16 <- list_census_vectors("CA16")
census_vectors21 <- list_census_vectors("CA21")

("CA21")

#write_csv(census_vectors21, "census_vectors21.csv")
```


```{r}

# Return data only
census_data <- get_census(dataset='CA16', regions=list(CMA="59933"), vectors=c("v_CA16_404","v_CA16_409","v_CA16_410"), level='CSD')


census_data_21 <- get_census(dataset='CA21', regions=list(PR="59"), level='CT')

```


## Regions

```{r}

# filters: 
#   "^" is equivalent to "starts with"
#   "$" is equivalent to "ends with"
#   "." is a single character wild card

census_regions %>%
  filter(grepl("^59", region) )

census_regions %>%
  filter(grepl("^59", region),
         level == "CMA")

```


## Data sets

### Income


```{r}
# Find the variables with "income" in the details

vector_income <- search_census_vectors("income", dataset = 'CA16')

vector_income



```

v_CA16_2397
Median total income of households in 2015 ($)

v_CA16_2398
Median after-tax income of households in 2015 ($)

v_CA16_2525
In low income based on the Low-income measure, after tax (LIM-AT)

v_CA16_2540
Prevalence of low income based on the Low-income measure, after tax (LIM-AT) (%)

v_CA16_2555
In low income based on the Low-income cut-offs, after tax (LICO-AT)





And example: Victoria

```{r}

# filters: 
#   "^" is equivalent to "starts with"
#   "$" is equivalent to "ends with"
#   "." is a single character wild card

# Return data only
data_935_CT <- get_census(dataset='CA16', regions=list(CMA="59935"),
                          vectors=c("v_CA16_2397","v_CA16_2398"), labels = "short", level = 'CT')

data_935_DA <- get_census(dataset='CA16', regions=list(CMA="59935"),
                          vectors=c("v_CA16_2397","v_CA16_2398"), labels = "short", level='DA')

head(data_935_DA)

```

add percentiles

```{r}

head(data_935_CT)

data_935_CT <- data_935_CT %>%
  mutate(med_pretax_pctl = percent_rank(`v_CA16_2397`),
         med_aftertax_pctl = percent_rank(`v_CA16_2398`))

data_935_DA <- data_935_DA %>%
  mutate(med_pretax_pctl = percent_rank(`v_CA16_2397`),
         med_aftertax_pctl = percent_rank(`v_CA16_2398`))


# use purrr


mtcars %>%
 select(., mpg, wt) %>% 
  map(., quantile, probs = c(0.10, 0.90), na.rm=TRUE) %>%
  as.data.frame(.) %>%
  `rownames<-`(c("10%", "90%"))

data_935_CT %>%
  select(., v_CA16_2397, v_CA16_2398) %>% 
  map(., quantile, probs = c(seq(0.0, 1.0, by = 0.1)), na.rm=TRUE) %>% 
  as.data.frame(.) %>% 
      `rownames<-`(c(seq(0.0, 1.0, by = 0.1)))

data_935_CT %>%
  select(., v_CA16_2397, v_CA16_2398) %>% 
  map(., quantile, probs = c(seq(0.0, 1.0, by = 0.2)), na.rm=TRUE) %>% 
  as.data.frame(.) %>% 
      `rownames<-`(c(seq(0.0, 1.0, by = 0.2)))

data_935_CT %>%
  select(., v_CA16_2397, v_CA16_2398) %>% 
  map(., quantile, probs = c(seq(0.0, 1.0, by = 0.25)), na.rm=TRUE) %>% 
  as.data.frame(.) %>% 
      `rownames<-`(c(seq(0.0, 1.0, by = 0.25)))


```


#### repeat for B.C.



```{r}
# Return data only
data_income_median_59_CT <- get_census(dataset='CA16', regions=list(PR="59"),
                          vectors=c("v_CA16_2397","v_CA16_2398"), labels = "short", level='CT')

data_income_median_59_DA <- get_census(dataset='CA16', regions=list(PR="59"),
                          vectors=c("v_CA16_2397","v_CA16_2398"), labels = "short", level='DA')

head(data_income_median_59_DA)


data_income_median_59_CT <- data_income_median_59_CT %>%
  mutate(med_pretax_pctl = percent_rank(`v_CA16_2397`),
         med_aftertax_pctl = percent_rank(`v_CA16_2398`))

data_income_median_59_DA <- data_income_median_59_DA %>%
  mutate(med_pretax_pctl = percent_rank(`v_CA16_2397`),
         med_aftertax_pctl = percent_rank(`v_CA16_2398`))

here()

write_csv(data_income_median_59_CT, 
          here("cancensus", "data", 
               "data_income_median_59_CT.csv"))


write_csv(data_income_median_59_DA, 
          here("cancensus", "data",
               "data_income_median_59_DA.csv"))



```




### Shelter to income ratio (STIR)

NOTE: add counts of owner and tenant households

```{r}

shelter_cost_vars <- c("v_CA16_4886", "v_CA16_4887", "v_CA16_4888", "v_CA16_4892", "v_CA16_4899")

shelter_cost_vars_list <- vector_income %>%
  filter(vector %in% shelter_cost_vars)

print(shelter_cost_vars_list)
write_csv(shelter_cost_vars_list, 
          here("cancensus", "data",
               "shelter_cost_vars_list.csv"))

# Return data only
data_sheltercost_935_CT <- get_census(dataset='CA16', regions=list(CMA="59935"),
                          vectors = shelter_cost_vars, labels = "short", level='CT')

data_sheltercost_935_DA <- get_census(dataset='CA16', regions=list(CMA="59935"),
                          vectors = shelter_cost_vars, labels = "short", level='DA')


```



#### repeat for B.C.

```{r}

#shelter_cost_vars <- c("v_CA16_4886", "v_CA16_4887", "v_CA16_4888", "v_CA16_4892", "v_CA16_4899")

#vector_income %>%
#  filter(vector %in% shelter_cost_vars)


# Return data only
data_sheltercost_59_CT <- get_census(dataset='CA16', regions=list(PR="59"),
                          vectors = shelter_cost_vars, labels = "short", level='CT')

data_sheltercost_59_DA <- get_census(dataset='CA16', regions=list(PR="59"),
                          vectors = shelter_cost_vars, labels = "short", level='DA')



# add percentiles 

data_sheltercost_59_CT <- data_sheltercost_59_CT %>%
  mutate(stir30_total_pctl = percent_rank(`v_CA16_4888`),
         stir30_owner_pctl = percent_rank(`v_CA16_4892`),
         stir30_tenant_pctl = percent_rank(`v_CA16_4899`))

data_sheltercost_59_DA <- data_sheltercost_59_DA %>%
  mutate(stir30_total_pctl = percent_rank(`v_CA16_4888`),
         stir30_owner_pctl = percent_rank(`v_CA16_4892`),
         stir30_tenant_pctl = percent_rank(`v_CA16_4899`))


# write as csv


write_csv(data_sheltercost_59_CT, 
          here("cancensus", "data",
               "data_sheltercost_59_CT.csv"))


write_csv(data_sheltercost_59_DA, 
                    here("cancensus", "data",
                         "data_sheltercost_59_DA.csv"))


```





Other variables

education
http://www12.statcan.gc.ca/census-recensement/2016/ref/dict/pop038-eng.cfm

housing - adequacy and suitability
http://www12.statcan.gc.ca/census-recensement/2016/ref/98-501/98-501-x2016007-eng.cfm

income - components of income
http://www12.statcan.gc.ca/census-recensement/2016/ref/98-501/98-501-x2016006-eng.cfm

income - all the low income
Low-income concepts
http://www12.statcan.gc.ca/census-recensement/2016/ref/98-501/98-501-x2016006-eng.cfm


households
http://www12.statcan.gc.ca/census-recensement/2016/ref/98-501/98-501-x2016004-eng.cfm
esp. persons living alone 
marital status

mobility-a-rama
http://www12.statcan.gc.ca/census-recensement/2016/ref/98-501/98-501-x2016014-eng.cfm

ethnicity
immigration past 5




map DA / CT to LHA



### Mobility


```{r}
# Find the variables with "mobility" in the details

vector_mobility <- search_census_vectors("mobility", dataset = 'CA16')

vector_mobility



```





### Disability

see "Activities of Daily Living" questions
https://www12.statcan.gc.ca/nhs-enm/2016/ref/questionnaires/questions-eng.cfm

```{r}
# Find the variables with "disability" in the details

vector_PWD <- search_census_vectors("difficulty", dataset = 'CA16')

vector_PWD

unique(vector_PWD$label)

```



# --- STOP HERE


 plagarized from Sasha
 

```{r}
censusMobility <- censusMobilityCma
censusMobility %<>% rename(
  # `Non-movers` = v_CA16_6695,
  Movers = v_CA16_6698,
  # `Non-migrants` = v_CA16_6701,
  Migrants = v_CA16_6704,
  `Internal migrants` = v_CA16_6707#,
  # `External migrants` = v_CA16_6716,
  # `Intraprovincial migrants` = v_CA16_6710,
  # `Interprovincial migrants` = v_CA16_6713,
)


vectorsMobility <- search_census_vectors(' Mobility status 1 year ago', "CA16", type = "Total") %>%
  child_census_vectors(leaves_only = FALSE)

for (censusLevel in c("CMA", "CD", "CSD", "CT", "DA")) {
  censusData <-
    get_census(
      "CA16",
      level = censusLevel,
      # regions = cdRegions,
      regions = regions,
      vectors = vectorsMobility %>% pull("vector"),
      use_cache = TRUE,
      labels = "short",
      geo_format = NA
    )
  censusData %<>%
    rename(
      `Non-movers` = v_CA16_6695,
      # Movers = v_CA16_6698,
      `Non-migrants` = v_CA16_6701,
      # Migrants = v_CA16_6704,
      # `Internal migrants` = v_CA16_6707,
      `External migrants` = v_CA16_6716,
      `Intraprovincial migrants` = v_CA16_6710,
      `Interprovincial migrants` = v_CA16_6713
    )
  censusData %<>% filter(Type == censusLevel)
  saveRDS(censusData, here::here("data", paste0("census2016-mobility-", censusLevel, ".rds")))
}

```


***

## CENSUS REFERENCE MATERIALS

Questions, 2016 https://www12.statcan.gc.ca/nhs-enm/2016/ref/questionnaires/questions-eng.cfm

-30-
