---
title: "Census: population by census tract"
subtitle: "population density 2021"
output: 
  html_document:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: inline
---

## `cancensus` - an introduction

** [`cancensus` github repo](https://github.com/mountainMath/cancensus) **


** [`cancensus` reference guide](https://mountainmath.github.io/cancensus/index.html)


> **NOTE**
> After an initial request to the cancensus server, all data is saved in the cache folder "cache_census". In other words, subsequent calls are to the local drive
> The cache location and other environment parameters are set in the file "cancensus_setup.R"


```{r setup}

# tidyverse
library(tidyverse)
library(here)

library(sf)
library(terra)  # for plotting population as dots

library(mapboxapi)

# cancensus
library(cancensus)
source(here::here("scripts", "cancensus_setup.R"))

# MNP plot format etc
source(here::here("scripts", "mnp_defaults_ggplot.R"))

# as_dot_density()
source(here::here("scripts", "as_dot_density.R"))

```



```{r}

# set API key (stored in text file "secretAPI.txt")
secretAPI <- readLines(here("scripts", "secretAPI.txt"))

options(cancensus.api_key = secretAPI)

```


```{r}
#set list of CSD names

CSD_master <- read_csv(
  here("geosuite", "2021_92-150-X_eng", "CSD.csv"),
  col_types =
    cols(CSDuid = col_character(),
         SACtype  = col_character())
)

```


***

Start exploring the data
(straight from [the `cancensus` reference guide](https://mountainmath.github.io/cancensus/index.html))




```{r}

# To view available Census datasets
census_datasets <- list_census_datasets()
census_datasets

```



####Regions

```{r}

# To view available named regions at different levels of Census hierarchy for the 2016 Census (for example)
#census_regions <- list_census_regions("CA16")
census_regions <- list_census_regions("CA21")


# filters: 
#   "^" is equivalent to "starts with"
#   "$" is equivalent to "ends with"
#   "." is a single character wild card

census_regions %>%
  filter(grepl("^59", region) )

census_regions %>%
  filter(grepl("^59", region),
         level == "CMA")

```


#### Population vectors 2021


```{r}

# To view available Census variables for the 2021 Census
# NOTE: there are 7,709 in total!!!

census_vectors <- list_census_vectors("CA21")
#write_csv(census_vectors, here("data_census", "census_vectors_CA21.csv"))

census_vectors

```


This list shows the vector numbers for the 2021 Census related to population


vector	type	label	units	parent_vector
v_CA21_1	Total	Population, 2021	Number	NA
v_CA21_2	Total	Population, 2016	Number	NA
v_CA21_3	Total	Population percentage change, 2016 to 2021	Number	NA
v_CA21_4	Total	Total private dwellings	Number	NA
v_CA21_5	Total	Private dwellings occupied by usual residents	Number	v_CA21_4
v_CA21_6	Total	Population density per square kilometre	Ratio	NA
v_CA21_7	Total	Land area in square kilometres	Number	NA
v_CA21_8	Total	Total - Age	Number	NA
v_CA21_9	Male	Total - Age	Number	NA
v_CA21_10	Female	Total - Age	Number	NA



```{r}
# list of vectors with the 2021 Census data 
vector_list <- c(
  "v_CA21_1", 
  "v_CA21_2", 
  "v_CA21_3",
  "v_CA21_4",
  "v_CA21_5", 
  "v_CA21_6", 
  "v_CA21_7",
  "v_CA21_8" 
  )


```



## Nanaimo CMA 938

```{r data_files_938}
# CENSUS SUB-DIVISIONS

cma_num <- "59938"

# Return data only
census_data <- get_census(dataset='CA21', regions=list(CMA=cma_num), vectors=vector_list, level='CSD')

# Return an sf-class data frame
census_data_sf <- get_census(dataset='CA21', regions=list(CMA=cma_num), vectors=vector_list, level='CSD', geo_format = "sf")

# Return a SpatialPolygonsDataFrame
census_data_sp <- get_census(dataset='CA21', regions=list(CMA=cma_num), vectors=vector_list, level='CSD', geo_format = "sp")
```


```{r data_files_938_CT}
# CENSUS TRACTS


cma_num <- "59938"


# Return data only
census_data <- get_census(dataset='CA21', regions=list(CMA=cma_num), vectors=vector_list, level='CT')

# Return an sf-class data frame
census_data_sf <- get_census(dataset='CA21', regions=list(CMA=cma_num), vectors=vector_list, level='CT', geo_format = "sf")

# Return a SpatialPolygonsDataFrame
census_data_sp <- get_census(dataset='CA21', regions=list(CMA=cma_num), vectors=vector_list, level='CT', geo_format = "sp")

```

mapping

```{r}



cma_num <- "59938"


# retrieve sf dataframe
nanaimo_cma <- get_census(dataset='CA21', regions=list(CMA=cma_num),
                         vectors=c(vector_list), level='CT', quiet = TRUE, 
                         geo_format = 'sf', labels = 'short')

nanaimo_cma <- nanaimo_cma %>% 
  left_join(CSD_master, by = c("CSD_UID" = "CSDuid"))
  
#write_rds(nanaimo_cma, here("data_census", "nanaimo_cma_2021.rds"))



```


```{r read_nanaimo_cma}

#nanaimo_cma <- read_rds(here("data_census", "nanaimo_cma_2021.rds"))

```


### pct pop change

v_CA21_3 == percent population change

```{r}

plot_data <- nanaimo_cma

ggplot(plot_data) + 
  geom_sf(aes(fill = v_CA21_3))

map_cma <- 
ggplot(plot_data) + 
  geom_sf(aes(fill = v_CA21_3)) +
  coord_sf(xlim = c(-124.1,-123.9),
           ylim = c(49.1, 49.3)) 

map_cma

```


viridis 1


```{r}

ggplot(nanaimo_cma) +
  geom_sf(aes(fill = v_CA21_3), colour = "grey") +
  scale_fill_gradient(
    low = "#eef0f2",
    high = "#0c3343",
    name = "population: percent change\n2016–2021", 
    labels = scales::comma
    ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = c(0.2, 0.75)
  ) +
  coord_sf(datum = NA) 

#  labs(title = "Median After-tax Income of Couple-Families with Children in 2015", subtitle = "Victoria CMA Census Tracts, 2016 Census")

#ggsave(here("plot_output", "cma938_2021_popchange.jpg"))

```

crop

```{r}

ggplot(nanaimo_cma) +
  geom_sf(aes(fill = v_CA21_3), colour = "grey") +
  # crop
  coord_sf(xlim = c(-124.1,-123.9),
           ylim = c(49.1, 49.3)) +
  
  scale_fill_gradient(
    low = "#eef0f2",
    high = "#0c3343",
     name = "percent population change, 2016–2021", 
    labels = scales::comma
   ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  ) 



```


### Nanaimo urban

#### net population change

```{r}


nanaimo_cma_urban <- nanaimo_cma |> 
  mutate(pop_change_net = Population - `Population 2016`) |> 
  filter(name != "0020.00")


ggplot(nanaimo_cma_urban) +
  geom_sf(aes(fill = pop_change_net), colour = "grey") +
  # crop
  coord_sf(xlim = c(-124.2,-123.7),
           ylim = c(49.00, 49.3)) +
  
  scale_fill_gradient2(
    low = "#ef5a27",
    high = "#0c3343",
     name = "Population:\nnet change\n2016–2021", 
    labels = scales::comma
   ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  ) 


#ggsave(here("plot_output", "cma938_2021_popchange_urban.jpg"))


```


#### population change: percent

```{r}
ggplot(nanaimo_cma_urban) +
  geom_sf(aes(fill = v_CA21_3), colour = "grey") +
  # crop
  coord_sf(xlim = c(-124.2,-123.7),
           ylim = c(49.00, 49.3)) +
  
  scale_fill_gradient2(
    low = "#ef5a27",
    high = "#0c3343",
     name = "Population:\npercent change\n2016–2021", 
    labels = scales::comma
   ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  ) 


#ggsave(here("plot_output", "cma938_2021_popchange_pct_urban.jpg"))


```


### population dots

From Kyle Walker

https://github.com/walkerke/map-challenge-2023/blob/main/scripts/day-27-dot.R



https://rspatial.github.io/terra/reference/dots.html

https://rspatial.org/index.html

https://bookdown.org/igisc/EnvDataSci/spatial-data-and-maps.html#spatvectors-in-terra


https://milospopovic.net/6-ways-to-map-population-with-r.r/




```{r}

nanaimo_dots <- as_dot_density(
  input_data = nanaimo_cma,
  value = "Population",
  values_per_dot = 100
)

```


```{r}

map_ct <- ggplot(nanaimo_cma) +
  geom_sf()


map_ct +
  geom_sf(
    mapping = aes(),
    data = nanaimo_dots,
    size = 0.01
    ) +
#  theme_void() +
#  scale_color_brewer(palette = "Set1") +
#  theme(legend.position = "bottom") +
  labs(
    #color = "Race / ethnicity",
       caption = "1 dot = 100 residents") 

```

#### Nanaimo urban


```{r}
map_ct_urban <- 
ggplot(nanaimo_cma_urban) +
  geom_sf() +
  # crop
  coord_sf(xlim = c(-124.2,-123.7),
           ylim = c(49.00, 49.3)) +
  
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  ) 

map_ct_urban
```


```{r}

people_per_dot <- 100

nanaimo_dots <- as_dot_density(
  input_data = nanaimo_cma_urban,
  value = "Population",
  values_per_dot = people_per_dot
)

map_ct_urban +
  geom_sf(
    mapping = aes(),
    data = nanaimo_dots,
    size = 0.01
    ) +
#  theme_void() +
#  scale_color_brewer(palette = "Set1") +
#  theme(legend.position = "bottom") +
  labs(
    #color = "Race / ethnicity",
       caption = glue::glue("1 dot = {people_per_dot} residents") 
  )

```

```{r}

people_per_dot <- 50

nanaimo_dots <- as_dot_density(
  input_data = nanaimo_cma_urban,
  value = "Population",
  values_per_dot = people_per_dot
)

map_ct_urban +
  geom_sf(mapping = aes(),
          data = nanaimo_dots,
          size = 0.01) +
  #  theme_void() +
  #  scale_color_brewer(palette = "Set1") +
  #  theme(legend.position = "bottom") +
  labs(
    title = "Nanaimo, British Columbia, Canada",
    subtitle = "Population, 2021 Census",
    caption = glue::glue(
      "1 dot = {people_per_dot} residents\nSource: Statistics Canada, 2021 Census"
    )
  )


#ggsave(here("plot_output", "cma938_2021_popdensity_urban.jpg"))


```


#### net increase

```{r}

people_per_dot <- 50

nanaimo_dots <- as_dot_density(
  input_data = nanaimo_cma_urban,
  value = "pop_change_net",
  values_per_dot = people_per_dot
)

map_ct_urban +
  geom_sf(mapping = aes(),
          data = nanaimo_dots,
          size = 0.01) +
  #  theme_void() +
  #  scale_color_brewer(palette = "Set1") +
  #  theme(legend.position = "bottom") +
  labs(
    title = "Nanaimo, British Columbia, Canada",
    subtitle = "Population change, 2021 Census",
    caption = glue::glue(
      "1 dot = {people_per_dot} new residents\nSource: Statistics Canada, 2021 Census"
    )
  )


#ggsave(here("plot_output", "cma938_2021_popdensity_growth_urban.jpg"))


```



## basemap


  coord_sf(xlim = c(-124.2,-123.7),
           ylim = c(49.00, 49.3)) +


```{r}
library(basemaps)
data(ext)
# or use draw_ext() to interactively draw an extent yourself

# view all available maps
get_maptypes()

# set defaults for the basemap
set_defaults(map_service = "osm", map_type = "topographic")

# load and return basemap map as class of choice, e.g. as image using magick:
basemap_magick(ext)

ext


```

### create bounding box

https://gis.stackexchange.com/questions/403977/sf-create-polygon-from-minimum-x-and-y-coordinates

```{r}

lon = c(-123.8155, -124.2002)
lat = c(49.1350,49.2632)


nanaimo_bb <- c(
  left = -123.9, right = -124.1000,
  bottom = 49.1350, top = 49.2632
  )
nanaimo_bb <- make_bbox(lon, lat)

```



Error in match.arg(maptype) : 
'arg' should be one of “stamen_terrain”, “stamen_toner”, “stamen_toner_lite”, “stamen_watercolor”, “alidade_smooth”, “alidade_smooth_dark”, “outdoors”, “stamen_terrain_background”, “stamen_toner_background”, “stamen_terrain_labels”, “stamen_terrain_lines”, “stamen_toner_labels”, “stamen_toner_lines”


```{r}
library(ggmap)

# Stadia maps API key
stadiaAPI <- readLines(here("scripts", "stadiaAPI.txt"))

ggmap::register_stadiamaps(stadiaAPI, write = FALSE)
#
#nanaimo_base_map <- ggmap::get_stadiamap(
#  bbox = nanaimo_bb,
#  maptype = "alidade_smooth",
#  zoom = 12
#)
#
#nanaimo_base_map
#
ggmap(nanaimo_basemap)
  

```

```{r}
ggmap(nanaimo_basemap) +
  geom_sf(mapping = aes(), inherit.aes = FALSE,
          data = nanaimo_dots,
          size = 0.01) +
  #  theme_void() +
  #  scale_color_brewer(palette = "Set1") +
  #  theme(legend.position = "bottom") +
  labs(
    title = "Nanaimo, British Columbia, Canada",
    subtitle = "Population, 2021 Census",
    caption = glue::glue(
      "1 dot = {people_per_dot} residents\nSource: Statistics Canada, 2021 Census"
    )
  )

  

```

zooming in: https://www.r-bloggers.com/2019/04/zooming-in-on-maps-with-sf-and-ggplot2/


### -30-

