---
title: "census tract"
subtitle: "population 2021"
output: 
  html_notebook:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: inline
---

#`cancensus` - an introduction

** [`cancensus` github repo](https://github.com/mountainMath/cancensus) **


** [`cancensus` reference guide](https://mountainmath.github.io/cancensus/index.html)





```{r setup}

# tidyverse
library(tidyverse)
library(here)

# cancensus

# (This is not necessary - `cancensus` is on CRAN now!)
#devtools::install_github("mountainMath/cancensus")

library(cancensus)


# cancensus
source(here::here("scripts", "cancensus_setup.R"))

# MNP plot format etc
source(here::here("scripts", "ggplot_mnp_defaults.R"))

```



```{r}

# set API key (stored in text file "secretAPI.txt")
secretAPI <- readLines(here("scripts", "secretAPI.txt"))

options(cancensus.api_key = secretAPI)

```


***

Start exploring the data
(straight from [the `cancensus` reference guide](https://mountainmath.github.io/cancensus/index.html))




```{r}

# To view available Census datasets
census_datasets <- list_census_datasets()
census_datasets

# To view available named regions at different levels of Census hierarchy for the 2016 Census (for example)
#census_regions <- list_census_regions("CA16")
census_regions <- list_census_regions("CA21")

# To view available Census variables for the 2021 Census
census_vectors <- list_census_vectors("CA21")
#write_csv(census_vectors, here("data_census", "census_vectors_CA21.csv"))
```

### Population vectors 2021

vector	type	label	units	parent_vector
v_CA21_1	Total	Population, 2021	Number	NA
v_CA21_2	Total	Population, 2016	Number	NA
v_CA21_3	Total	Population percentage change, 2016 to 2021	Number	NA
v_CA21_4	Total	Total private dwellings	Number	NA
v_CA21_5	Total	Private dwellings occupied by usual residents	Number	v_CA21_4
v_CA21_6	Total	Population density per square kilometre	Ratio	NA
v_CA21_7	Total	Land area in square kilometres	Number	NA
v_CA21_8	Total	Total - Age	Number	NA
v_CA21_9	Male	Total - Age	Number	NA
v_CA21_10	Female	Total - Age	Number	NA



```{r}
# list of vectors with the 2021 Census data for total by single year ages, 5 to 12
vector_list <- c(
  "v_CA21_1", 
  "v_CA21_2", 
  "v_CA21_3",
  "v_CA21_4",
  "v_CA21_5", 
  "v_CA21_6", 
  "v_CA21_7",
  "v_CA21_8" 
  )


```


### data: Vancouver CMA 933

```{r data_files}

# CENSUS SUB-DIVISIONS

# Return data only
census_data <- get_census(dataset='CA21', regions=list(CMA="59933"), vectors=vector_list, level='CSD')

# Return an sf-class data frame
census_data_sf <- get_census(dataset='CA21', regions=list(CMA="59933"), vectors=vector_list, level='CSD', geo_format = "sf")

# Return a SpatialPolygonsDataFrame
census_data_sp <- get_census(dataset='CA21', regions=list(CMA="59933"), vectors=vector_list, level='CSD', geo_format = "sp")



# CENSUS TRACTS

# Return data only
census_data <- get_census(dataset='CA21', regions=list(CMA="59933"), vectors=vector_list, level='CT')

# Return an sf-class data frame
census_data_sf <- get_census(dataset='CA21', regions=list(CMA="59933"), vectors=vector_list, level='CT', geo_format = "sf")

# Return a SpatialPolygonsDataFrame
census_data_sp <- get_census(dataset='CA21', regions=list(CMA="59933"), vectors=vector_list, level='CT', geo_format = "sp")

```


## Regions

```{r}

# filters: 
#   "^" is equivalent to "starts with"
#   "$" is equivalent to "ends with"
#   "." is a single character wild card

census_regions %>%
  filter(grepl("^59", region) )

census_regions %>%
  filter(grepl("^59", region),
         level == "CMA")

```


## Data sets


### CT

```{r}

# filters: 
#   "^" is equivalent to "starts with"
#   "$" is equivalent to "ends with"
#   "." is a single character wild card

# Return data only
# Return data only
data_CA21_933_CSD <- get_census(dataset = 'CA21', 
                          regions = list(CMA = "59933"),
                          vectors = vector_list, 
                          labels = "short", 
                          level = 'CSD')

data_CA21_933_CT <- get_census(dataset = 'CA21', 
                          regions = list(CMA = "59933"),
                          vectors = vector_list, 
                          labels = "short", 
                          level = 'CT')


head(data_CA21_933_CT)

#write_csv(data_CA21_933_CT, here("cancensus", "data", "data_CA21_933_CT_vCA16_2397.csv"))

data_CA21_933_DA <- get_census(dataset='CA21', regions=list(CMA="59933"),
                          vectors = vector_list, labels = "short", level='DA')


```




### Census tracts

```{r westshore_definition}

# unique CSD (municipality) values

csd_933 <- as_tibble(unique(data_CA21_933_CT$CSD_UID))

# join CSD names

CSD_master <- read_csv(
  here("geosuite", "2021_92-150-X_eng", "CSD.csv"),
  col_types =
    cols(CSDuid = col_character(),
         SACtype  = col_character())
)


data_CA21_933_CT <- data_CA21_933_CT %>% 
  left_join(CSD_master, by = c("CSD_UID" = "CSDuid"))




```



```{r read_write_midstream}

#write_rds(data_CA21_933_CT, here("data_census", "data_CA21_933_CT_population.rds"))
#write_csv(data_CA21_933_CT, here("data_census", "data_CA21_933_CT_population.csv"))

#data_CA21_933_CT <- read_rds(here("data_census", "data_CA21_933_CT_population.rds"))


```


### summary stats


### mapping


```{r}
# retrieve sf dataframe
vancouver_cma <- get_census(dataset='CA21', regions=list(CMA="59933"),
                         vectors=c(vector_list), level='CT', quiet = TRUE, 
                         geo_format = 'sf', labels = 'short')

vancouver_cma <- vancouver_cma %>% 
  left_join(CSD_master, by = c("CSD_UID" = "CSDuid"))
  
#write_rds(vancouver_cma, here("data_census", "vancouver_cma_2021.rds"))



```

### read Vancouver 933

```{r}

vancouver_cma <- read_rds(here("data_census", "vancouver_cma_2021.rds"))

```


```{r}

plot_data <- vancouver_cma
fill_var <- "v_CA21_6"

ggplot(plot_data) + 
  geom_sf(aes(fill = v_CA21_6))

map_cma <- 
ggplot(plot_data) + 
  geom_sf(aes(fill = v_CA21_6)) +
  coord_sf(
    xlim = c(-122.66, -123.26),
    ylim = c(48.95, 49.3))

map_cma

```


viridis 1

```{r}

ggplot(vancouver_cma) +
  geom_sf(aes(fill = v_CA21_6), colour = "grey") +
  scale_fill_gradient(
    low = "#eef0f2",
    high = "#0c3343",
    name = "population density, 2021", 
    labels = scales::comma
    ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = c(0.2, 0.75)
  ) +
  coord_sf(datum = NA) 

#  labs(title = "Median After-tax Income of Couple-Families with Children in 2015", subtitle = "Victoria CMA Census Tracts, 2016 Census")

# ggsave(here("plot_output", "map01_income.png"))

```

crop

```{r}

ggplot(vancouver_cma) +
  geom_sf(aes(fill = v_CA21_6), colour = "grey") +
  # crop
  coord_sf(xlim = c(-123.18,-123.05),
           ylim = c(49.32, 49.25)) +
  
#  scale_fill_viridis_c("Median HH Income", labels = scales::dollar) +
  scale_fill_gradient(
    low = "#eef0f2",
    high = "#0c3343",
     name = "persons per sq.km, 2021", 
    labels = scales::comma
   ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  ) 

#+
  #coord_sf(datum = NA) +
#  labs(title = "Number of children ages 5–12", subtitle = "Victoria CMA Census Subdivisions, 2021 Census")

```

zooming in: https://www.r-bloggers.com/2019/04/zooming-in-on-maps-with-sf-and-ggplot2/


#### Langley 2021


```{r Langley_population_density}


vancouver_cma %>% 
  distinct(CSDname, CSDtype)

langley_dm <- vancouver_cma %>% 
  filter(CSDname == "Langley" & CSDtype == "DM") %>% 
  rename(
    "population_2021" = Population,
    "population_2016" = `Population 2016`
    ) 

langley_dm <- langley_dm %>% 
  mutate(
    pop_density_2021 = population_2021 / `Shape Area`,
    pop_density_2016 = population_2016 / `Shape Area`
  )


min(langley_dm$v_CA21_6)
max(langley_dm$v_CA21_6)

```



```{r}

# plot  
ch_pop_density_2021 <- ggplot(langley_dm) +
  geom_sf(aes(fill = v_CA21_6), colour = "grey") +
  #
  scale_fill_gradient(
    low = "#eef0f2",
    high = "#0c3343",
    limits = c(80, 4000),
     name = "persons per sq.km", 
    labels = scales::comma
   ) +
  theme_light() +
  theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  ) +
  labs(
    title = "Township of Langley (District Municipality)",
    caption = "Source: Statistics Canada, Census 2021",
    subtitle = "Population density, 2021"
  )

#ggsave(here("outputs", "Langley_DM_pop_density_CT_2021.png"))
#ggsave(here("outputs", "Langley_DM_pop_density_CT_2021.jpg"))

ch_pop_density_2021

```


#### Langley 2016

```{r Langley_population_density_2016}

#ls.str(langley_dm)  

#min(langley_dm$pop_density_2016)
#max(langley_dm$pop_density_2016)


# plot  
ch_pop_density_2016 <-
ggplot(langley_dm) +
  geom_sf(aes(fill = pop_density_2016), colour = "grey") +
#  geom_sf(aes(fill = pop_density_2016), show.legend = FALSE) +

  scale_fill_gradient(
    low = "#eef0f2",
    high = "#0c3343",
    limits = c(80, 4000),
     name = "persons per sq.km", 
    labels = scales::comma
   ) +
  theme_light() +
  theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  ) +
  labs(
    title = "Township of Langley (District Municipality)",
    caption = "Source: Statistics Canada, Census 2016",
    subtitle = "Population density, 2016"
  )

#ggsave(here("outputs", "Langley_DM_pop_density_CT_2016.png"))
#ggsave(here("outputs", "Langley_DM_pop_density_CT_2016.jpg"))

ch_pop_density_2016

```

#### Langley 2016 & 2021

```{r}

gridExtra::grid.arrange(
  ch_pop_density_2016, 
  ch_pop_density_2021, 
  ncol = 2
  )

```


#### facet for years

```{r}

#ls.str(langley_dm)

langley_dm_long <- langley_dm %>% 
  select(GeoUID, geometry, pop_density_2021, pop_density_2016) %>% 
  pivot_longer(
    cols = c(pop_density_2021, pop_density_2016),
    names_to = "census_year",
    names_prefix = "pop_density_",
    values_to = "pop_density"
  )

langley_dm_long

```



```{r}

# plot  
ch_pop_density_both <- ggplot(langley_dm_long) +
  geom_sf(aes(fill = pop_density), colour = "grey") +
  facet_wrap(~ census_year) +
  #
  scale_fill_gradient(
    low = "#eef0f2",
    high = "#0c3343",
    limits = c(80, 4000),
     name = "persons per sq.km", 
    labels = scales::comma
   ) +
  theme_light() +
  theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  ) +
  labs(
    title = "Township of Langley (District Municipality)",
    caption = "Source: Statistics Canada, Census 2021",
    subtitle = "Population density, 2016 & 2021"
  )

#ggsave(here("outputs", "Langley_DM_pop_density_CT__2016_2021.png"))
#ggsave(here("outputs", "Langley_DM_pop_density_CT__2016_2021.jpg"))

ch_pop_density_both

```

