---
title: "Census: mobility by Census Subdivisions"
subtitle: "2021"
output: 
  html_document:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: inline
---

## Census Mobility

**Defintions etc**

https://www12.statcan.gc.ca/census-recensement/2021/ref/98-500/010/98-500-x2021010-eng.cfm

Mobility status
Mobility status refers to the status of a person with regard to their place of residence on Census Day (May 11, 2021) in relation to their place of residence on the same date 1 or 5 years earlier.

Persons who, on Census Day, lived in the same residence as they did on the same date 1 year or 5 years earlier are referred to as non-movers, and persons who, on Census Day, did not live in the same residence as they did 1 year or 5 years earlier are referred to as movers.

Movers include non-migrants and migrants.

Non-migrants are movers who lived in the same census subdivision on Census Day as they did on the same date 1 year or 5 years earlier.

Migrants include internal migrants and external migrants.

Internal migrants include migrants who lived in Canada 1 year or 5 years ago. This includes persons who moved to a different city, town, township, village, municipality or Indian reserve within Canada. Greater detail can be provided for internal migrants by identifying whether they crossed a provincial boundary or the limits of a large municipality, specifically a census metropolitan area (CMA) or census agglomeration (CA).

External migrants include migrants who did not live in Canada 1 year or 5 years ago.


## `cancensus` - an introduction

** [`cancensus` github repo](https://github.com/mountainMath/cancensus) **


** [`cancensus` reference guide](https://mountainmath.github.io/cancensus/index.html)


> **NOTE**
> After an initial request to the cancensus server, all data is saved in the cache folder "cache_census". In other words, subsequent calls are to the local drive
> The cache location and other environment parameters are set in the file "cancensus_setup.R"


```{r setup}

# tidyverse
library(tidyverse)
library(here)

library(sf)
library(terra)  # for plotting population as dots

#library(mapboxapi)

# cancensus
library(cancensus)
source(here::here("scripts", "cancensus_setup.R"))

# MNP plot format etc
source(here::here("scripts", "mnp_defaults_ggplot.R"))

# as_dot_density()
source(here::here("scripts", "as_dot_density.R"))

```



```{r}

# set API key (stored in text file "secretAPI.txt")
secretAPI <- readLines(here("scripts", "secretAPI.txt"))

options(cancensus.api_key = secretAPI)

```


```{r}
#set list of CSD names

CSD_master <- read_csv(
  here("geosuite", "2021_92-150-X_eng", "CSD.csv"),
  col_types =
    cols(CSDuid = col_character(),
         SACtype  = col_character())
)

```


***

Start exploring the data
(straight from [the `cancensus` reference guide](https://mountainmath.github.io/cancensus/index.html))




```{r}

# To view available Census datasets
census_datasets <- list_census_datasets()
census_datasets

```



#### Regions

```{r}

# To view available named regions at different levels of Census hierarchy for the 2016 Census (for example)
#census_regions <- list_census_regions("CA16")
census_regions <- list_census_regions("CA21")


# filters: 
#   "^" is equivalent to "starts with"
#   "$" is equivalent to "ends with"
#   "." is a single character wild card

census_regions %>%
  filter(grepl("^59", region) )

census_regions %>%
  filter(grepl("^59", region),
         level == "CSD")

```


#### Household vectors 2021


```{r}

# To view available Census variables for the 2021 Census
# NOTE: there are 7,709 in total!!!

census_vectors <- list_census_vectors("CA21")
#write_csv(census_vectors, here("data_census", "census_vectors_CA21.csv"))

census_vectors

```

See the file "census_vectors_CA21.csv" for a complete list


This list shows the vector numbers for the 2021 Census related to households

* check the CSV file for parent_vector relationships


v_CA21_5745	Total	Total - Mobility status 1 year ago
v_CA21_5746	Male	Total - Mobility status 1 year ago
v_CA21_5747	Female	Total - Mobility status 1 year ago
v_CA21_5748	Total	Non-movers
v_CA21_5749	Male	Non-movers
v_CA21_5750	Female	Non-movers
v_CA21_5751	Total	Movers
v_CA21_5752	Male	Movers
v_CA21_5753	Female	Movers
v_CA21_5754	Total	Non-migrants
v_CA21_5755	Male	Non-migrants
v_CA21_5756	Female	Non-migrants
v_CA21_5757	Total	Migrants
v_CA21_5758	Male	Migrants
v_CA21_5759	Female	Migrants
v_CA21_5760	Total	Internal migrants
v_CA21_5761	Male	Internal migrants
v_CA21_5762	Female	Internal migrants
v_CA21_5763	Total	Intraprovincial migrants
v_CA21_5764	Male	Intraprovincial migrants
v_CA21_5765	Female	Intraprovincial migrants
v_CA21_5766	Total	Interprovincial migrants
v_CA21_5767	Male	Interprovincial migrants
v_CA21_5768	Female	Interprovincial migrants
v_CA21_5769	Total	External migrants
v_CA21_5770	Male	External migrants
v_CA21_5771	Female	External migrants
v_CA21_5772	Total	Total - Mobility status 5 years ago
v_CA21_5773	Male	Total - Mobility status 5 years ago
v_CA21_5774	Female	Total - Mobility status 5 years ago
v_CA21_5775	Total	Non-movers
v_CA21_5776	Male	Non-movers
v_CA21_5777	Female	Non-movers
v_CA21_5778	Total	Movers
v_CA21_5779	Male	Movers
v_CA21_5780	Female	Movers
v_CA21_5781	Total	Non-migrants
v_CA21_5782	Male	Non-migrants
v_CA21_5783	Female	Non-migrants
v_CA21_5784	Total	Migrants
v_CA21_5785	Male	Migrants
v_CA21_5786	Female	Migrants
v_CA21_5787	Total	Internal migrants
v_CA21_5788	Male	Internal migrants
v_CA21_5789	Female	Internal migrants
v_CA21_5790	Total	Intraprovincial migrants
v_CA21_5791	Male	Intraprovincial migrants
v_CA21_5792	Female	Intraprovincial migrants
v_CA21_5793	Total	Interprovincial migrants
v_CA21_5794	Male	Interprovincial migrants
v_CA21_5795	Female	Interprovincial migrants
v_CA21_5796	Total	External migrants
v_CA21_5797	Male	External migrants
v_CA21_5798	Female	External migrants



Five year, total

```{r}
# list of vectors with the 2021 Census data 
vector_list <- c(
'v_CA21_5772', #	Total	Total - Mobility status 5 years ago
'v_CA21_5775',	# Total	Non-movers
'v_CA21_5778',	# Total	Movers
'v_CA21_5781',	# Total	Non-migrants
'v_CA21_5784',	# Total	Migrants
'v_CA21_5787',	# Total	Internal migrants
'v_CA21_5790',	# Total	Intraprovincial migrants
'v_CA21_5793',	# Total	Interprovincial migrants
'v_CA21_5796' 	# Total	External migrants
)


```



## Comox Valley CD 5926

CD 5926 == Comox Valley Regional District



```{r data_files_938}
# CENSUS SUB-DIVISIONS

cma_num <- "5926"

# Return data only
census_data <- get_census(dataset='CA21', regions=list(CMA=cma_num), vectors=vector_list, level='CSD') |> 
  janitor::clean_names()

# Return an sf-class data frame
census_data_sf <- get_census(dataset='CA21', regions=list(CMA=cma_num), vectors=vector_list, level='CSD', geo_format = "sf") |> 
  janitor::clean_names()

# Return a SpatialPolygonsDataFrame
census_data_sp <- get_census(dataset='CA21', regions=list(CMA=cma_num), vectors=vector_list, level='CSD', geo_format = "sp") |> 
  janitor::clean_names()

```


check the hierarchy

```{r}

comox <- census_data |> 
  filter(
    geo_uid == "5926005"
  ) |> 
  pivot_longer(
    cols = v_ca21_5772_total_mobility_status_5_years_ago:v_ca21_5796_external_migrants,
    names_to = "census_variable",
    values_to = "count_households"
  )

write_csv(comox, "comox.csv")

```



create custom household structure hierarchy

```{r}
mobility_grp <- census_data |> 
  select(
    geo_uid,
    region_name,
    "non-movers" = v_ca21_5775_non_movers,
    "non-migrants" = v_ca21_5781_non_migrants,
    "intra-provincial migrants" = v_ca21_5790_intraprovincial_migrants,
    "inter-provincial migrants" = v_ca21_5793_interprovincial_migrants,
    "external migrants" = v_ca21_5796_external_migrants
  ) |> 
  pivot_longer(
    -c(geo_uid:region_name),
    names_to = "mobility_type",
    values_to = "person_count"
  ) |> 
  # filter out IRI -- no data
  filter(region_name != "Pentledge 2 (IRI)")

mobility_grp <- mobility_grp |> 
  group_by(region_name) |> 
  mutate(pct_region = round((person_count / sum(person_count) * 100), 1))
```


```{r}
# set factor order

mobility_grp <- mobility_grp |> 
  mutate(mobility_type = as_factor(mobility_type))

levels(mobility_grp$mobility_type)

 

```



### plot: mobility CD

```{r}
# count
ggplot(mobility_grp, aes(x = reorder(mobility_type, desc(mobility_type)), y = person_count)) +
    geom_col() + 
  coord_flip()
```


```{r}
# fancy it up


pop_range_breaks <- seq(0, 40000, by = 10000)


ggplot(mobility_grp, aes(x = reorder(mobility_type, desc(mobility_type)), y = person_count)) +
  geom_col(fill = "#0c3343") +
  coord_flip() +
  mnp_chart_theme +
  theme(
    # remove all gridlines
    panel.grid = element_blank()
  ) +
    labs(caption = "Source: Statistics Canada, Census 2021",
         x = NULL,
         y = "number of persons") +
  scale_y_continuous(
    breaks  = pop_range_breaks,
    limits = c(min(pop_range_breaks), max(pop_range_breaks)),
#    labels = abs(pop_range_breaks),
    labels = scales::label_comma()
  )  +
  # add vertical white "gridlines"
  geom_hline(yintercept = pop_range_breaks, col = "white")


ggsave(here::here("plot_output/comox_cd_mobility_type.jpg")
#       width = 9,
#       height = 6
  )

```



by CSD: municipality only


```{r}
mobility_grp_muni <- mobility_grp |> 
  filter(region_name %in% c("Comox (T)", "Courtenay (CY)", "Cumberland (VL)"))

levels(mobility_grp_muni$region_name)
```


```{r}
# clever function to put minor ticks without a label
# https://stackoverflow.com/questions/34533472/insert-blanks-into-a-vector-for-e-g-minor-tick-labels-in-r

every_nth <- function(x, nth, empty = TRUE, inverse = FALSE) 
  {
  if (!inverse) {
    if(empty) {
      x[1:nth == 1] <- ""
      x
      } else {
        x[1:nth != 1]
        }
    } else {
      if(empty) {
        x[1:nth != 1] <- ""
        x
        } else {
          x[1:nth == 1]
        }
    }
}

```


```{r}
# fancy it up



pop_range_breaks <- seq(0, 100, by = 10)


ggplot(mobility_grp_muni,
       aes(x = reorder(mobility_type, desc(mobility_type)),
           y = pct_region)) +
  geom_col(fill = "#0c3343") +
  facet_wrap( ~ region_name) +
  coord_flip() +
  mnp_chart_theme +
  theme(# remove all gridlines
    panel.grid = element_blank()) +
  labs(caption = "Source: Statistics Canada, Census 2021",
       x = NULL,
       y = "percent of persons") +
  scale_y_continuous(
    breaks  = pop_range_breaks,
 #   minor_breaks = pop_range_breaks / 2,
    limits = c(min(pop_range_breaks), max(pop_range_breaks)),
    #    labels = abs(pop_range_breaks),
    labels = every_nth(pop_range_breaks, 2, inverse = TRUE))  +
  # add vertical white "gridlines"
  geom_hline(yintercept = (pop_range_breaks), col = "white") +
  # add bar value labels
  geom_text(
    #data = mobility_grp_muni,
    aes(
#      label = pct_region
      label = round(pct_region, 0)
      ),
    colour = "#0c3343",
    hjust = -.5)


ggsave(here::here("plot_output/comox_csd_mobility_muni.jpg")
#       width = 9,
#       height = 6
  )


```



### -30-

