---
title: "Census: households by Census Subdivisions"
subtitle: "2021"
output: 
  html_document:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: inline
---

## CENSUS HOUSEHOLDS

**Defintions etc**

https://www23.statcan.gc.ca/imdb/p3Var.pl?Function=Unit&Id=32746

https://www23.statcan.gc.ca/imdb/p3Var.pl?Function=DEC&Id=251053#:~:text=Census%20family%20households%20are%20those,not%20constitute%20a%20census%20family.


"Other Census Households" combo group:

Code |	Category
One-census-family household without additional persons | One-census-family household without additional persons
Multigenerational household | Multigenerational household
Other census family household | Other census family household
One-person household | One-person household
Two-or-more-person non-census-family household | Two-or-more-person non-census-family household


## `cancensus` - an introduction

** [`cancensus` github repo](https://github.com/mountainMath/cancensus) **


** [`cancensus` reference guide](https://mountainmath.github.io/cancensus/index.html)


> **NOTE**
> After an initial request to the cancensus server, all data is saved in the cache folder "cache_census". In other words, subsequent calls are to the local drive
> The cache location and other environment parameters are set in the file "cancensus_setup.R"


```{r setup}

# tidyverse
library(tidyverse)
library(here)

library(sf)
library(terra)  # for plotting population as dots

#library(mapboxapi)

# cancensus
library(cancensus)
source(here::here("scripts", "cancensus_setup.R"))

# MNP plot format etc
source(here::here("scripts", "mnp_defaults_ggplot.R"))

# as_dot_density()
source(here::here("scripts", "as_dot_density.R"))

```



```{r}

# set API key (stored in text file "secretAPI.txt")
secretAPI <- readLines(here("scripts", "secretAPI.txt"))

options(cancensus.api_key = secretAPI)

```


```{r}
#set list of CSD names

CSD_master <- read_csv(
  here("geosuite", "2021_92-150-X_eng", "CSD.csv"),
  col_types =
    cols(CSDuid = col_character(),
         SACtype  = col_character())
)

```


***

Start exploring the data
(straight from [the `cancensus` reference guide](https://mountainmath.github.io/cancensus/index.html))




```{r}

# To view available Census datasets
census_datasets <- list_census_datasets()
census_datasets

```



#### Regions

```{r}

# To view available named regions at different levels of Census hierarchy for the 2016 Census (for example)
#census_regions <- list_census_regions("CA16")
census_regions <- list_census_regions("CA21")


# filters: 
#   "^" is equivalent to "starts with"
#   "$" is equivalent to "ends with"
#   "." is a single character wild card

census_regions %>%
  filter(grepl("^59", region) )

census_regions %>%
  filter(grepl("^59", region),
         level == "CSD")

```


#### Household vectors 2021


```{r}

# To view available Census variables for the 2021 Census
# NOTE: there are 7,709 in total!!!

census_vectors <- list_census_vectors("CA21")
#write_csv(census_vectors, here("data_census", "census_vectors_CA21.csv"))

census_vectors

```

See the file "census_vectors_CA21.csv" for a complete list


This list shows the vector numbers for the 2021 Census related to households

* check the CSV file for parent_vector relationships


vector	type	label	units	parent_vector
v_CA21_443	Total	Private households by household size
v_CA21_444	Total	1 person
v_CA21_445	Total	2 persons
v_CA21_446	Total	3 persons
v_CA21_447	Total	4 persons
v_CA21_448	Total	5 or more persons
v_CA21_449	Total	Number of persons in private households
v_CA21_450	Male	Number of persons in private households
v_CA21_451	Female	Number of persons in private households
v_CA21_452	Total	Average household size
v_CA21_510	Total	Persons in private households
v_CA21_511	Male	Persons in private households
v_CA21_512	Female	Persons in private households
v_CA21_513	Total	Persons in census families
v_CA21_514	Male	Persons in census families
v_CA21_515	Female	Persons in census families
v_CA21_516	Total	Married spouses or common-law partners
v_CA21_517	Male	Married spouses or common-law partners
v_CA21_518	Female	Married spouses or common-law partners
v_CA21_519	Total	Parents in one-parent families
v_CA21_520	Male	Parents in one-parent families
v_CA21_521	Female	Parents in one-parent families
v_CA21_522	Total	Children
v_CA21_523	Male	Children
v_CA21_524	Female	Children
v_CA21_525	Total	In a two-parent family
v_CA21_526	Male	In a two-parent family
v_CA21_527	Female	In a two-parent family
v_CA21_528	Total	In a one-parent family
v_CA21_529	Male	In a one-parent family
v_CA21_530	Female	In a one-parent family
v_CA21_531	Total	Persons not in census families in private households
v_CA21_532	Male	Persons not in census families in private households
v_CA21_533	Female	Persons not in census families in private households
v_CA21_534	Total	Living alone
v_CA21_535	Male	Living alone
v_CA21_536	Female	Living alone
v_CA21_537	Total	Living with other relatives
v_CA21_538	Male	Living with other relatives
v_CA21_539	Female	Living with other relatives
v_CA21_540	Total	Living with non-relatives only
v_CA21_541	Male	Living with non-relatives only
v_CA21_542	Female	Living with non-relatives only
v_CA21_543	Total	Household type
v_CA21_544	Total	One-census-family households without additional persons
v_CA21_545	Total	Couple-family households
v_CA21_546	Total	With children
v_CA21_547	Total	Without children
v_CA21_548	Total	One-parent-family households
v_CA21_549	Total	Multigenerational households
v_CA21_550	Total	Multiple-census-family households
v_CA21_551	Total	One-census-family households with additional persons
v_CA21_552	Total	Two-or-more-person non-census-family households
v_CA21_553	Total	One-person households



```{r}
# list of vectors with the 2021 Census data 
vector_list <- c(
  "v_CA21_543", 
  "v_CA21_544", 
  "v_CA21_545", 
  "v_CA21_546", 
  "v_CA21_547", 
  "v_CA21_548", 
  "v_CA21_549", 
  "v_CA21_550", 
  "v_CA21_551", 
  "v_CA21_552", 
  "v_CA21_553"
  )


```



## Comox Valley CD 5926

CD 5926 == Comox Valley Regional District



```{r data_files_938}
# CENSUS SUB-DIVISIONS

cma_num <- "5926"

# Return data only
census_data <- get_census(dataset='CA21', regions=list(CMA=cma_num), vectors=vector_list, level='CSD') |> 
  janitor::clean_names()

# Return an sf-class data frame
census_data_sf <- get_census(dataset='CA21', regions=list(CMA=cma_num), vectors=vector_list, level='CSD', geo_format = "sf") |> 
  janitor::clean_names()

# Return a SpatialPolygonsDataFrame
census_data_sp <- get_census(dataset='CA21', regions=list(CMA=cma_num), vectors=vector_list, level='CSD', geo_format = "sp") |> 
  janitor::clean_names()

```


check the hierarchy

```{r}

comox <- census_data |> 
  filter(
    geo_uid == "5926005"
  ) |> 
  pivot_longer(
    cols = v_ca21_543_household_type:v_ca21_553_one_person_households,
    names_to = "census_variable",
    values_to = "count_households"
  )

write_csv(comox, "comox.csv")

```



create custom household structure hierarchy

```{r}

household_grp <- census_data |> 
  rowwise() |> 
  mutate(
    `other census households` = sum(c_across(v_ca21_549_multigenerational_households:v_ca21_552_two_or_more_person_non_census_family_households))
  ) |> 
  select(
    geo_uid,
    region_name,
    "total households" = v_ca21_543_household_type,
    "couple family with children" = v_ca21_546_with_children,
    "couple family w/o children" = v_ca21_547_without_children,
    "one parent" = v_ca21_548_one_parent_family_households,
    "one person" = v_ca21_553_one_person_households,
    "other census households"
  ) |> 
  pivot_longer(
    -c(geo_uid:region_name),
    names_to = "household_type",
    values_to = "household_count"
  )

# Comox Valley - merge IRI areas
household_grp <- household_grp |> 
  mutate(region_name_merge = case_when(
    str_detect(region_name, "IRI") ~ "Comox 1 & Pentledge 2 (IRI)",
    TRUE ~ region_name
  )) |> 
  group_by(region_name_merge, household_type) |> 
  summarise(household_count = sum(household_count, na.rm = TRUE)) |> 
  rename("region_name" = region_name_merge)

household_grp

```

```{r}
household_grp_2 <- household_grp |> 
  filter(household_type != "total households") |> 
  group_by(region_name) |> 
  mutate(household_pct = household_count / sum(household_count) * 100) 


household_grp_2$household_type |> 
  fct_relevel(
    "couple family w/o children",
    "couple family with children",
    "one parent",
    "one person",
    "other census households"
    ) |> 
  levels()

```


### plot: household type

```{r}
# count
ggplot(household_grp_2, aes(x = household_type, y = household_count)) +
    geom_col() + 
  coord_flip()
```


```{r}
# fancy it up



pop_range_breaks <- seq(0, 12000, by = 2000)


ggplot(household_grp_2, aes(x = household_type, y = household_count)) +
  geom_col(fill = "#0c3343") +
  coord_flip() +
  mnp_chart_theme +
  theme(
    # remove all gridlines
    panel.grid = element_blank()
  ) +
    labs(caption = "Source: Statistics Canada, Census 2021",
         x = NULL,
         y = "number of households") +
   # flip the order of the categories
  scale_x_discrete(limits = rev) +
  scale_y_continuous(
    breaks  = pop_range_breaks,
    limits = c(min(pop_range_breaks), max(pop_range_breaks)),
#    labels = abs(pop_range_breaks),
    labels = scales::label_comma()
  )  +
  # add vertical white "gridlines"
  geom_hline(yintercept = pop_range_breaks, col = "white")

ggsave(
  here::here("plot_output/comox_cd_household_type.jpg")
#  width = 13.333,
#  height = 7.5
)


```



by CSD

```{r}
# count
ggplot(household_grp_2, aes(x = region_name, y = household_count, fill = household_type)) +
    geom_col() + 
  coord_flip() +
  scale_fill_brewer("blues", direction = -1)


# percent
ggplot(household_grp_2, aes(x = region_name, y = household_pct, fill = household_type)) +
    geom_col() +
  coord_flip() +
  scale_fill_brewer("blues", direction = -1)


```




### -30-

