---
title: "population data retreval"
author: "Martin Monkman"
date: "2023-03-02"
output: html_document
---

```{r setup}
# packages
library(tidyverse)
library(readxl)
# cansim
# reference page: https://mountainmath.github.io/cansim/index.html
library(cansim) 


# create list of Vancouver Island CDs numbers
vi_cd_list <- c("17", "19", "21", "23", "24", "26", "27", "43", "45")

five_year_bins <- c(
  "0 to 4 years",
  "5 to 9 years",
  "10 to 14 years",
  "15 to 19 years",
  "20 to 24 years",
  "25 to 29 years",
  "30 to 34 years",
  "35 to 39 years",
  "40 to 44 years",
  "45 to 49 years",
  "50 to 54 years",
  "55 to 59 years",
  "60 to 64 years",
  "65 to 69 years",
  "70 to 74 years",
  "75 to 79 years",
  "80 to 84 years",
  "85 to 89 years",
  "90 to 94 years",
  "95 to 99 years",
  "100 years and over"  
)


```


## Census 2021 - single year


Age (in single years), average age and median age and gender: Census metropolitan areas, census agglomerations and census subdivisions1
Frequency: Occasional

Table: 98-10-0021-01

Release date: 2022-04-27

Geography: Census subdivision, Census metropolitan area, Census agglomeration

Universe: Total Population, including institutional residents, 2021 and 2016 censuses — 100% data

Variable List: Age (in single years), average age and median age (128), Census year (2), Gender (3a)


How to cite: Statistics Canada. Table 98-10-0021-01  Age (in single years), average age and median age and gender: Census metropolitan areas, census agglomerations and census subdivisions
DOI: https://doi.org/10.25318/9810002101-eng


https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=9810002101

_NOTE: 98-10-0021-01 does NOT include Tofino and Ucluelet, perhaps due to boundary changes between the Census years_

Age (in single years), average age and median age and gender: Canada, provinces and territories, census divisions, census subdivisions and dissemination areas
Frequency: Occasional

Table: 98-10-0023-01

Release date: 2022-04-27

Geography: Canada, Province or territory, Census division, Census subdivision, Dissemination area

Universe: Total population, including institutional residents, 2021 Census — 100% data

Variable List: Age (in single years), average age and median age (128), Gender (3)

https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=9810002301

How to cite: Statistics Canada. Table 98-10-0023-01  Age (in single years), average age and median age and gender: Canada, provinces and territories, census divisions, census subdivisions and dissemination areas
DOI: https://doi.org/10.25318/9810002301-eng


```{r}


df_pop_csd_detail <- read_csv("data_statcan/98100021.csv") |> 
  janitor::clean_names()


```

```{r}
df_pop_csd_bc <- df_pop_csd_detail |> 
  filter(str_detect(dguid, "2021A000559")) |> 
# strip geo into component parts
  # geocode with parenthesis
  mutate(geo_code = str_match(geo, "\\(\\d{7}\\)")) |> 
  mutate(geo_code = str_remove(geo_code, "^\\(")) |> 
  mutate(geo_code = str_remove(geo_code, "\\)$")) |> 
  # cd and csd numbers separately
  mutate(geo_cd = as.character(str_sub(dguid, start = 12, end = 13)))  |> 
  mutate(geo_csd = as.character(str_sub(dguid, start = 14, end = 16)))  |> 
  # place name
  mutate(geo_name = str_match(geo, "^.*\\(\\d")) |> 
  mutate(geo_name = str_remove(geo_name, " \\(\\d$")) |> 
  # geo type
  mutate(geo_type = str_match(geo, " \\D+$")) |> 
  mutate(geo_type = str_trim(geo_type))

df_pop_csd_bc

#write_rds(df_pop_csd_bc, "data_statcan/df_pop_csd_bc.rds")
```

```{r}

#df_pop_csd_bc <- read_rds("data_statcan/df_pop_csd_bc.rds")
  
df_pop_csd_bc_5yr <- df_pop_csd_bc |> 
  filter(age_in_single_years_average_age_and_median_age_128 %in%
           five_year_bins) |> 
  rename(
    age_grp = age_in_single_years_average_age_and_median_age_128,
    pop_total = gender_3a_total_gender_1,
    pop_men = gender_3a_men_2,
    pop_women = gender_3a_women_3,
    census_year = census_year_2
) |> 
  mutate(age_grp = as_factor(age_grp))

levels(df_pop_csd_bc_5yr$age_grp)

#write_rds(df_pop_csd_bc_5yr, "data_statcan/df_pop_csd_bc_5yr.rds")

```


## BC 5 year

```{r}

# Vancouver Island CDs

df_pop_csd_vi_5yr <- df_pop_csd_bc_5yr |> 
  select(
    census_year,
    geo_name,
    geo_type,
    geo_cd,
    geo_csd,
    age_grp,
    pop_total,
    pop_men,
    pop_women
  ) |> 
  filter(
    geo_cd %in% vi_cd_list,
    census_year == "2021")

df_pop_csd_vi_5yr

# check

df_pop_csd_vi_5yr |> 
  summarise(
    total_pop = sum(pop_total, na.rm = TRUE)
  )

```

```{r}

# calculate gender as pct of total, by geo
df_pop_csd_vi_5yr <- df_pop_csd_vi_5yr |> 
  group_by(geo_name, geo_type) |> 
  mutate(
    total_pop_csd = sum(pop_total, na.rm = TRUE),
    total_men_csd = sum(pop_men, na.rm = TRUE),
    total_women_csd = sum(pop_women, na.rm = TRUE)
    ) 

  # checks -- doesn't add up due to random rounding
#  mutate(
#    total_total = total_men_csd + total_women_csd
#  ) |> 
#  mutate(
#    total_check = total_pop_csd - total_total
#  ) |> 
#  select(geo, total_check)
#

df_pop_csd_vi_5yr <- df_pop_csd_vi_5yr |> 
  mutate(
    pct_men = pop_men / total_pop_csd * 100,
    pct_women = pop_women / total_pop_csd * 100
  ) |> 
  mutate(pct_age_grp = pct_men + pct_women) |> 
  ungroup()

# check
df_pop_csd_vi_5yr |> 
  group_by(geo_name) |> 
  summarise(total_pct_100 = sum(pct_age_grp))

df_pop_csd_vi_5yr



```

### Vancouver Island total

```{r}
df_pop_csd_vi_5yr_total <- df_pop_csd_vi_5yr |> 
  group_by(age_grp) |> 
  summarise(
    total_age_grp = sum(pop_total, na.rm = TRUE),
    total_men = sum(pop_men, na.rm = TRUE),
    total_women = sum(pop_women, na.rm = TRUE)
    ) |> 
  ungroup()

df_pop_csd_vi_5yr_total <- df_pop_csd_vi_5yr_total |> 
  mutate(total_population = sum(total_age_grp)) |> 
    mutate(
    pct_men = total_men / total_population * 100,
    pct_women = total_women / total_population * 100
  )

df_pop_csd_vi_5yr_total

```

```{r}

# prep for plot
df_pl_vi_pyramid <- 
  df_pop_csd_vi_5yr_total |> 
  select(age_grp, pct_men, pct_women) |> 
  mutate(pct_men = -pct_men) |> 
  pivot_longer(-age_grp, names_to = "gender", values_to = "pct_of_pop") 
  
  # and plot
pl_pyramid_vi <-  
  ggplot(df_pl_vi_pyramid, aes(x = pct_of_pop,
           y = age_grp,
           colour = gender
           )) + 
  geom_col(fill = NA) +
  theme_light()

pl_pyramid_vi

```

fix scale numbering

```{r}

pop_range_breaks <- seq(-5, 5, by = 1)

pl_pyramid_vi +
   scale_x_continuous(breaks  = pop_range_breaks,
                       labels = abs(pop_range_breaks))  

```



### pop pyramid: csd

prep for plot

```{r}
df_pop_pyramid_csd <- df_pop_csd_vi_5yr |> 
  select(-contains("total"),
         -contains("pct_age_grp")) |> 
  # set value of men to negative
  mutate(
    pop_men = -pop_men,
    pct_men = -pct_men
  ) |> 
  pivot_longer(-c(census_year:age_grp), names_to = "gender", values_to = "pop_2021") 

df_pop_pyramid_csd
```

```{r}

# list of place names in "geo"
community_names <- unique(df_pop_pyramid_csd$geo_name)
community_names


```


### plot community

```{r}
community_name <- "Campbell River"

df_pop_pyramid_community <- df_pop_pyramid_csd |> 
  # filter for pop or pct
  filter(str_detect(gender, "pop")) |> 
  # filter for geography
  filter(geo_name == community_name) 
  
  # now plot
pl_pyramid_csd <- 
ggplot(
  df_pop_pyramid_community,
  aes(x = pop_2021,
           y = age_grp,
           fill = gender)) + 
  geom_col()

# pl_pyramid_csd

# fix scale numbering

pop_range <- range(df_pop_pyramid_community$pop_2021)
pop_range

pop_range_breaks <- pretty(pop_range)

pl_pyramid_csd +
   scale_x_continuous(breaks  = pop_range_breaks,
                       labels = abs(pop_range_breaks))  

```



### overlay VI region




```{r}
pl_pyramid_csd <- 
df_pop_pyramid_csd |> 
  # filter for pop or pct
  filter(str_detect(gender, "pct")) |> 
  # filter for geography
  filter(geo_name == community_name) |> 
  
  # now plot
ggplot(aes(x = pop_2021,
           y = age_grp,
           fill = gender)) + 
  geom_col()

pl_pyramid_csd

# overlay region
pl_pyramid_csd <- pl_pyramid_csd +
  geom_col(
    aes(
      x = df_pl_vi_pyramid$pct_of_pop, 
      y = df_pl_vi_pyramid$age_grp,
      colour = gender 
      ),
    colour = "black",
    size = 1,
    fill = NA 
    )


# fix scale numbering

pop_range_breaks <- seq(-10, 10, by = 2)

pl_pyramid_csd +
   scale_x_continuous(breaks  = pop_range_breaks,
                       labels = abs(pop_range_breaks))  


```

