---
title: "population data retreval"
author: "Martin Monkman"
date: "2023-03-02"
output: html_document
---

```{r setup}
# packages
library(tidyverse)
library(readxl)
# cansim
# reference page: https://mountainmath.github.io/cansim/index.html
library(cansim) 


# create list of Vancouver Island CDs numbers
vi_cd_list <- c("17", "19", "21", "23", "24", "26", "27", "43", "45")

five_year_bins <- c(
  "0 to 4 years",
  "5 to 9 years",
  "10 to 14 years",
  "15 to 19 years",
  "20 to 24 years",
  "25 to 29 years",
  "30 to 34 years",
  "35 to 39 years",
  "40 to 44 years",
  "45 to 49 years",
  "50 to 54 years",
  "55 to 59 years",
  "60 to 64 years",
  "65 to 69 years",
  "70 to 74 years",
  "75 to 79 years",
  "80 to 84 years",
  "85 to 89 years",
  "90 to 94 years",
  "95 to 99 years",
  "100 years and over"  
)


```


## Data

### by CD

"CD" == "Census Division" == BC Regional District (the other provinces have different administrative regions that align with CDs)


Population estimates, July 1, by census division, 2016 boundaries1, 2, 3, 4, 5, 6
Frequency: Annual

Table: 17-10-0139-01

Release date: 2023-01-11

Geography: Canada, Province or territory, Census division

```{r}

#pop_cd <- get_cansim("17-10-0139-01")

write_rds(pop_cd, "pop_cd.rds")
```

#### apply filters

```{r}
# BC total population
pop_cd %>% 
  filter(GEO == "British Columbia") %>% 
  filter(Sex == "Both sexes" &
           `Age group` == "All ages")

# Vancouver Island CDs 
pop_cd_vi <- pop_cd %>%
  #filter(str_detect(GeoUID, "59\\d\\d")) %>%
  filter(str_detect(GeoUID, "^59")) %>%
  filter(Sex == "Both sexes" &
           `Age group` == "All ages") %>% 
  
  # filter for Vancouver Island
  # create separate variable with just the CD number
  mutate(geo_cd = as.character(str_sub(GeoUID, start = 3, end = 4)))  %>% 
  filter(geo_cd %in% vi_cd_list)
  
pop_cd_vi

```

clean up Regional District names 

* split into regional district name and province columns

```{r}
# split GEO into regional district and province
pop_cd_vi <- pop_cd_vi %>% 
  tidyr::separate_wider_delim(GEO, delim = ", ", names = c("regional_district", "province"))

pop_cd_vi

```


```{r}
# remove unnecessary columns, put into wide format
pop_cd_vi_wide <- pop_cd_vi %>% 
  select(
    REF_DATE, 
    regional_district, 
    GeoUID,
    VALUE
    ) %>% 
  pivot_wider(names_from = REF_DATE, values_from = VALUE)

pop_cd_vi_wide

```



### by CSD

CDs are divided into CSDs

"CSD" == "Census Sub-Division" == Municipality, formal Indian Reserves and other First Nation lands, & other unincorporated rural areas (in BC, known as Regional District Electoral Areas)

#### historical trends

Population estimates, July 1, by census subdivision, 2016 boundaries1, 2, 3, 4, 5, 6
Frequency: Annual

Table: 17-10-0142-01

Release date: 2023-01-11

Geography: Canada, Province or territory, Census subdivision

https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1710014201

```{r}

#pop_csd <- get_cansim("17-10-0142-01")
#write_rds(pop_csd, "pop_csd.rds")

pop_csd <- read_rds("pop_csd.rds")
```

filter for BC & Vancouver Island


```{r}


pop_csd_bc <- pop_csd %>% 
  filter(str_detect(DGUID, "2016A000559")) %>% 
  mutate(geo_cd = as.character(str_sub(DGUID, start = 12, end = 13)))  %>% 
  mutate(geo_csd = as.character(str_sub(DGUID, start = 14, end = 16)))  
  
pop_csd_bc

# Vancouver Island CDs

pop_csd_bc %>% 
  filter(geo_cd %in% vi_cd_list)

```



#### Census 2021 - single year


Age (in single years), average age and median age and gender: Census metropolitan areas, census agglomerations and census subdivisions1
Frequency: Occasional

Table: 98-10-0021-01

Release date: 2022-04-27

Geography: Census subdivision, Census metropolitan area, Census agglomeration

Universe: Total Population, including institutional residents, 2021 and 2016 censuses — 100% data

Variable List: Age (in single years), average age and median age (128), Census year (2), Gender (3a)


How to cite: Statistics Canada. Table 98-10-0021-01  Age (in single years), average age and median age and gender: Census metropolitan areas, census agglomerations and census subdivisions
DOI: https://doi.org/10.25318/9810002101-eng


https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=9810002101

```{r}


df_pop_csd_detail <- read_csv("data_statcan/98100021.csv") |> 
  janitor::clean_names()

df_pop_csd_detail |> 
  distinct(ref_date)

```

```{r}
df_pop_csd_bc <- df_pop_csd_detail |> 
  filter(str_detect(dguid, "2021A000559")) %>% 
  mutate(geo_cd = as.character(str_sub(dguid, start = 12, end = 13)))  %>% 
  mutate(geo_csd = as.character(str_sub(dguid, start = 14, end = 16)))  
  
df_pop_csd_bc_5yr <- df_pop_csd_bc |> 
  filter(age_in_single_years_average_age_and_median_age_128 %in%
           five_year_bins) |> 
  rename(
    age_grp = age_in_single_years_average_age_and_median_age_128,
    pop_total = gender_3a_total_gender_1,
    pop_men = gender_3a_men_2,
    pop_women = gender_3a_women_3
) |> 
  mutate(age_grp = as_factor(age_grp))

levels(df_pop_csd_bc_5yr$age_grp)

```


```{r}

# Vancouver Island CDs

df_pop_csd_vi_5yr <- df_pop_csd_bc_5yr %>% 
  filter(geo_cd %in% vi_cd_list) |> 
  select(
    geo,
    geo_cd,
    geo_csd,
    age_grp,
    pop_total,
    pop_men,
    pop_women
  ) 

df_pop_csd_vi_5yr

# calculate gender as pct of total, by geo
df_pop_csd_vi_5yr <- df_pop_csd_vi_5yr |> 
  group_by(geo) |> 
  mutate(
    total_pop_csd = sum(pop_total, na.rm = TRUE),
    total_men_csd = sum(pop_men, na.rm = TRUE),
    total_women_csd = sum(pop_women, na.rm = TRUE)
    ) 

  # checks -- doesn't add up due to random rounding
#  mutate(
#    total_total = total_men_csd + total_women_csd
#  ) |> 
#  mutate(
#    total_check = total_pop_csd - total_total
#  ) |> 
#  select(geo, total_check)
#

df_pop_csd_vi_5yr <- df_pop_csd_vi_5yr |> 
  mutate(
    pct_men = pop_men / total_pop_csd * 100,
    pct_women = pop_women / total_pop_csd * 100
  ) |> 
  mutate(pct_age_grp = pct_men + pct_women) |> 
  ungroup()

# check
df_pop_csd_vi_5yr |> 
  group_by(geo) |> 
  summarise(total_pct_100 = sum(pct_age_grp))

df_pop_csd_vi_5yr



```


```{r}
df_pop_csd_vi_5yr_total <- df_pop_csd_vi_5yr |> 
  group_by(age_grp) |> 
  summarise(
    total_pop = sum(pop_total, na.rm = TRUE),
    total_men = sum(pop_men, na.rm = TRUE),
    total_women = sum(pop_women, na.rm = TRUE)
    )

df_pop_csd_vi_5yr_total |> 
  mutate(total_population = sum(total_pop)) |> 
    mutate(
    pct_men = total_men / total_men * 100,
    pct_women = total_women / total_women * 100
  )
```


  # now plot
ggplot(aes(x = pop_2021,
           y = age_grp,
           colour = gender)) + 
  geom_col(fill = "white")


### pop pyramid: csd

prep for plot

```{r}
df_pop_pyramid <- df_pop_csd_vi_5yr |> 
  select(-contains("total"),
         -contains("pct_age_grp")) |> 
  # set value of men to negative
  mutate(
    pop_men = -pop_men,
    pct_men = -pct_men
  ) |> 
  pivot_longer(-c(geo:age_grp), names_to = "gender", values_to = "pop_2021") 

```


```{r}
df_pop_pyramid |> 
  # filter for pop or pct
  filter(str_detect(gender, "pop")) |> 
  # filter for geography
  filter(str_detect(geo, "Oak Bay")) |> 
  
  # now plot
ggplot(aes(x = pop_2021,
           y = age_grp,
           colour = gender)) + 
  geom_col(fill = "white")

```

