---
title: "Vancouver Island population"
author: "Martin Monkman"
date: "2023-06-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# package setup
library(tidyverse)

# cansim
# reference page: https://mountainmath.github.io/cansim/index.html
library(cansim) 

```



## Data

## by CD

"CD" == "Census Division" == BC Regional District (the other provinces have different administrative regions that align with CDs)


Vancouver Island CDs:

```{r}

# create list of Vancouver Island CDs numbers
vi_cd_list <- c("17", "19", "21", "23", "24", "26", "27", "43", "45")
```



Population estimates, July 1, by census division, 2016 boundaries1, 2, 3, 4, 5, 6
Frequency: Annual

Table: 17-10-0139-01

Release date: 2023-01-11

Geography: Canada, Province or territory, Census division

https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1710013901


```{r}

### --- DATA CALL

pop_cd <- get_cansim("17-10-0139-01")

```


#### apply filters


```{r}


# filter for Vancouver Island CDs 
pop_cd_vi <- pop_cd |>
  # filter for BC, total population
  filter(str_detect(GeoUID, "^59")) |>
  filter(Sex == "Both sexes" &
           `Age group` == "All ages") |> 
  # filter for Vancouver Island
  # - create separate variable with just the CD number
  mutate(geo_cd = as.character(str_sub(GeoUID, start = 3, end = 4)))  |> 
  filter(geo_cd %in% vi_cd_list)

```

clean up Regional District names 

* split into regional district name and province columns


```{r}
# split GEO into regional district and province
pop_cd_vi <- pop_cd_vi |> 
  tidyr::separate_wider_delim(GEO, delim = ", ", names = c("regional_district", "province"))
```




```{r}

# remove unnecessary columns, reshape into wide format
pop_cd_vi_wide <- pop_cd_vi |> 
  select(
    REF_DATE, 
    regional_district, 
    GeoUID,
    VALUE
  ) |> 
  pivot_wider(names_from = REF_DATE, values_from = VALUE)


pop_cd_vi_wide

```


#### alternate path

Now we have the appropriate table filtered, it's possible to call only the specific variables (vectors) that we need.

This is MUCH faster--it doesn't have to pull all the rest of Canada.

The downside is that it doesn't have the label, which we have to get with `get_cansim_vector_info()`

```{r}

# make list of vector numbers

lst_vector_cd_vi <- pop_cd_vi |> 
  distinct(VECTOR) |> 
  pull(VECTOR)
  
lst_vector_cd_vi

```

```{r}

pop_vi_vector <- get_cansim_vector(lst_vector_cd_vi)

```

Get labels; join to main table

```{r}
tbl_vi_cd <- get_cansim_vector_info(lst_vector_cd_vi) |> 
  select(VECTOR, title_en) |> 
  separate(title_en, sep = ";", into = c("cd", "sex", "age")) |> 
  mutate(cd = str_remove(cd, ", British Columbia"))
  
tbl_vi_cd

```

```{r}

pop_vi_vector |> 
  left_join(tbl_vi_cd, by = "VECTOR")
```



## by CSD

CDs are divided into CSDs

"CSD" == "Census Sub-Division" == Municipality, formal Indian Reserves and other First Nation lands, & other unincorporated rural areas (in BC, known as Regional District Electoral Areas)


Population estimates, July 1, by census subdivision, 2016 boundaries1, 2, 3, 4, 5, 6
Frequency: Annual

Table: 17-10-0142-01

Release date: 2023-01-11

Geography: Canada, Province or territory, Census subdivision

https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1710014201

```{r}

pop_csd <- get_cansim("17-10-0142-01")

```

filter for BC & Vancouver Island


```{r}


pop_csd_bc <- pop_csd %>% 
  filter(str_detect(DGUID, "2016A000559")) %>% 
  mutate(geo_cd = as.character(str_sub(DGUID, start = 12, end = 13)))  %>% 
  mutate(geo_csd = as.character(str_sub(DGUID, start = 14, end = 16)))  
  
pop_csd_bc

# Vancouver Island CDs

pop_csd_bc %>% 
  filter(geo_cd %in% vi_cd_list)

```


